---
title: "Live demo exercises"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "On this page"
    toc-float: true
    toc-depth: 3
---

[Return to course materials page](https://cambiotraining.github.io/data-analysis-in-r-and-python/)

## Section setup

::: {.panel-tabset group="language"}
## R

```{r}
#| message: false
library(tidyverse)
```

## Python

:::

## DA1: Getting started

## DA2: Data & plotting

Live demo's.

### Live demo: exploring `infections`

::: {.panel-tabset group="language"}
## R

Read in the data:

```{r}
#| message: false
infections <- read_csv("../materials/data/infections.csv")
```


## Python
:::

#### Data structure

::: {.panel-tabset group="language"}
## R

Number of rows & columns:

```{r}
#| eval: false
nrow(infections)
```

```{r}
#| eval: false
ncol(infections)
```

It's good to look at the column attributes: what type of columns are we dealing with and is it what we expect?

```{r}
#| eval: false
summary(infections)
```

```{r}
#| eval: false
str(infections)
```

## Python
:::

#### Quality control checks

It's good to do some basic sanity / quality control checks. For example, if there are different categories in a column, do all the categories we expect show up or are there missing ones / misspelled etc.?

::: {.panel-tabset group="language"}
## R

For example, we can check the unique values in a column:

```{r}
unique(infections$infection_type)
```

We can count the number of missing values in the column `infection_type`.
You read the code "inside-out":

```{r}
sum(is.na(infections$infection_type))
```

## Python

:::

### Live demo: subsetting `infections`

::: {.panel-tabset group="language"}
## R

Let's select `patient_id`:

```{r}
#| eval: false
infections$patient_id
```

Or more than 1 column, by column name:

```{r}
infections[, c("patient_id", "systolic_pressure")]
```

Combine this with selecting only a subset of rows:

```{r}
infections[1:3, c("patient_id", "systolic_pressure")]
```

## Python
:::

## DA3: Manipulating data

## DA4: Organise and combine

