---
title: Manipulating columns
---

::: {.callout-tip}
#### Learning objectives

- Learn how to select and/or rename specific columns from a data frame
- Be able to create new columns and modify existing ones
:::


## Context

In the [tabular data](#tabular-data) section we learned to deal with, well, tabular data in the form of our `surveys` data set. This data set isn't huge, but sometimes we have many variables and we might only want to work with a subset of them. Or, we might want to create new columns based on existing data. In this section we'll cover how we can do this.

## Section setup {#setup_plotting}

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

We'll start this section with a new script named, for example, `05-manipulation.R`. If needed, add the following code to the top of your script and run it.

```{r}
#| message: false
# A collection of R packages designed for data science
library(tidyverse)

surveys <- read_csv("data/surveys.csv")
```

## Python

We'll start this section with a new script named, for example, `05-manipulation.py`. If needed, add the following code to the top of your script and run it.

```{python}
# A Python data analysis and manipulation tool
import pandas as pd

# Python equivalent of `ggplot2`
from plotnine import *

surveys = pd.read_csv("data/surveys.csv")
```

:::
:::

## Selecting columns

Let's remind ourselves to which columns we have in our `surveys` data set. After that, we'll start making some changes.

::: {.panel-tabset group="language"}
## R

```{r}
colnames(surveys)
```
 
## Python

```{python}
surveys.columns
```

:::

### Selecting individual columns

Let's say we wanted to select only the `record_id` and `year` columns. We've briefly done this when we looked at [subsetting rows and columns](#subset_rc).

::: {.panel-tabset group="language"}
## R

However, there is an alternative way of doing this using the `dplyr` package - which is part of `tidyverse`.
 
 We can use the `select()` function:
 
```{r}
select(surveys, record_id, year)
```
 
Using the base R syntax, this is equivalent to `surveys[, c("record_id", "year")]`. Notice that with the `select()` function (and generally with `dplyr` functions) we didn’t need to quote " the column names. This is because the first input to the function is the table name, and so everything after is assumed to be column names of that table.

## Python

The way we need to specify this is by giving a list of column names `["record_id", "year"]` and subsetting the surveys data set with this.

The way we subset is with `surveys[ ]`, so we end up with double square brackets:

```{python}
surveys[["record_id", "year"]]
```

:::

### Selecting with helper functions

::: {.panel-tabset group="language"}
## R

The `select()` function becomes particularly useful when we combine it with other helper functions. For example, this code will select all the columns where the column name contains the string (text) `"_id"`:

```{r}
# returns all columns where the column name contains the text "_id"
select(surveys, contains("_id"))
```

## Python

The subsetting becomes a bit tedious when we're looking for patterns in the column names. Here, we can instead use the `.filter` attribute of the surveys data set, and look for a string (text) where the column name contains `"_id"`.

```{python}
surveys.filter(like = "_id")
```


:::

### Selecting a range of columns

Let's say we're interested in all the columns from `record_id` to `year`.

::: {.panel-tabset group="language"}
## R
In that case, we can use the `:` symbol.

```{r}
# returns all columns between and including record_id and year
select(surveys, record_id:year)
```

We can also combine this with the previous method:

```{r}
# returns all columns between and including record_id and year
# and all columns where the column name contains the text "_id"
select(surveys, record_id:year, contains("_id"))
```

## Python

In that case, we can use the `:` symbol, in combination with the `.loc` indexer.

```{python}
surveys.loc[:, "record_id":"year"]
```


:::

### Unselecting columns

Lastly, we can also **unselect columns**. This can be useful when you want most columns, apart from some.

::: {.panel-tabset group="language"}
## R

To do this, we use the `-` symbol before the column name.

```{r}
# returns all columns apart from record_id
select(surveys, -record_id)
```

## Python

To do this, we use the `.drop` attribute. Here, we only unselect one column, but we can easily extend this by providing a list of columns do the `column =` argument.

```{python}
surveys.drop(columns = "record_id")
```

:::

## Renaming and reshuffling columns

### Renaming columns

For example, we might want to change the `weight` column name to `weight_g`, to reflect that the values are in grams.

::: {.panel-tabset group="language"}
## R
We can use the `rename()` function to change a column name. We do this as follows:

```{r}
rename(surveys, weight_g = weight)
```

## Python

We can use the `.rename()` attribute of the `surveys` Pandas data frame:

```{python}
surveys.rename(columns = {'weight': 'weight_g'})
```

:::

### Reshuffling columns

It might be that you want to reorder/reshuffle a column. Here, the `year` column is our fourth variable. Let's say we'd want to  move this to the second position (after `record_id`).

::: {.panel-tabset group="language"}
## R

We can use the `relocate()` function to do this. The function has several arguments, starting with `.`, such as `.before =` or `.after =`. These allow you to specify where you want to reinsert the column.

```{r}

relocate(surveys, year, .after = record_id)
```


## Python

Unlike in R, there isn't a very clear, straightforward way of reinserting columns in a Pandas data frame. We could show you convoluted ways of doing so, but at this point that's just confusing. So, we'll leave you with a [link to a Stackoverflow solution](https://stackoverflow.com/questions/13148429/how-to-change-the-order-of-dataframe-columns).
:::

## Creating new columns
LO: creating new columns



## Summary

::: {.callout-tip}
#### Key points

- 
:::
