---
title: Manipulating columns
---

::: {.callout-tip}
#### Learning objectives

- Learn how to select and/or rename specific columns from a data frame
- Be able to create new columns and modify existing ones
:::


## Context
In [tabular data](#tabular-data) section we learned to deal with, well, tabular data in the form of our `surveys` data set. This data set isn't huge, but sometimes we have many variables and we might only want to work with a subset of them. Or, we might want to create new columns based on existing data. In this section we'll cover how we can do this.

## Section setup {#setup_plotting}

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

We'll start this section with a new script named, for example, `05-manipulation.R`. If needed, add the following code to the top of your script and run it.

```{r}
#| message: false
# A collection of R packages designed for data science
library(tidyverse)

surveys <- read_csv("data/surveys.csv")
```

## Python

We'll start this section with a new script named, for example, `05-manipulation.py`. If needed, add the following code to the top of your script and run it.

```{python}
# A Python data analysis and manipulation tool
import pandas as pd

# Python equivalent of `ggplot2`
from plotnine import *

surveys = pd.read_csv("data/surveys.csv")
```

:::
:::

## Manipulation of variables

### Selecting columns

Let's remind ourselves to which columns we have in our `surveys` data set. After that, we'll start making some changes.

::: {.panel-tabset group="language"}
## R

```{r}
colnames(surveys)
```
 
 Let's say we wanted to select only the `record_id` and `year` columns. We've briefly done this when we looked at [subsetting rows and columns](#subset_rc). However, there is an alternative way of doing this using the `dplyr` package - which is part of `tidyverse`.
 
 We can use the `select()` function:
 
```{r}
select(surveys, record_id, year)
```
 
Using the base R syntax, this is equivalent to `surveys[, c("record_id", "year")]`. Notice that with the `select()` function (and generally with `dplyr` functions) we didn’t need to quote " the column names. This is because the first input to the function is the table name, and so everything after is assumed to be column names of that table.

The `select()` function becomes particularly useful when we combine it with other helper functions. For example, this code will select all the columns where the column name contains the string (text) `"_id"`:

```{r}
# returns all columns where the column name contains the text "_id"
select(surveys, contains("_id"))
```


Or maybe we're interested in all the columns from `record_id` to `year`. In that case, we can use the `:` symbol.

```{r}
# returns all columns between and including record_id and year
select(surveys, record_id:year)
```

We can also combine these:

```{r}
# returns all columns between and including record_id and year
# and all columns where the column name contains the text "_id"
select(surveys, record_id:year, contains("_id"))
```

Lastly, we can also **unselect columns**. This can be useful when you want most columns, apart from some. To do this, we use the `-` symbol before the column name. For example:

```{r}
# returns all columns apart from record_id
select(surveys, -record_id)
```

## Python
:::

### Renaming and reshuffling columns
LO: renaming and reshuffling columns

### Creating new columns
LO: creating new columns



## Summary

::: {.callout-tip}
#### Key points

- 
:::
