---
title: Clean, style & arrange
---

::: {.callout-tip}
#### Learning objectives

- Learn how to deal with messy variables names.
- Be able to update `id` columns.
- Show confidence in fixing common encoding issues.
- Know how to tidy plot labels.
- Be able to change plot colours.
- Learn how to arrange composite plots (in `R`).
:::


## Context
Often data is in a messy state before you can work with it. So, it is useful to know when and how to make changes to your data.

## Section setup {#setup_grouped_operations}

::: {.panel-tabset group="language"}
## R

We'll continue this section with the script named `05_session`. If needed, add the following code to the top of your script and run it.

```{r}
#| message: false
# A collection of R packages designed for data science
library(tidyverse)

# A package for cleaning up data
library(janitor)

# A package for creating composite plots
library(patchwork)

messy_data <- read_csv("data/messy_data.csv")
surveys <- read_csv("data/surveys.csv")
plot_types <- read_csv("data/plots.csv")
surveys_left <- left_join(surveys, plot_types, by = "plot_id")
```

## Python

We'll continue this section with the Notebook named `05_session`. Add the following code to the first cell and run it.


```{python}
# A Python data analysis and manipulation tool
import pandas as pd

# Python equivalent of `ggplot2`
from plotnine import *

# A package for cleaning up data
import janitor

# A package for text wrapping
import textwrap

messy_data = pd.read_csv("data/messy_data.csv")
surveys = pd.read_csv("data/surveys.csv")
plot_types = pd.read_csv("data/plots.csv")
surveys_left = pd.merge(surveys, plot_types, how = "left", on = "plot_id")
```

:::

## Cleaning data

Perhaps it's not the most glamorous part of data analysis, but it is a very important one: cleaning data. If you need motivation for it, just think of the amount of time you can save yourself by recording your data consistently and correctly in the first place.

In the next few sections we'll go through some (very) common messy data issues you're likely to come across. We will revisit some of our previous data sets, but will mostly illustrate things using the `messy` data set. This data set has been synthesised for this exact purpose, so it's over-the-top bad. If you ever come across a real data set that looks like this, then a stern word with that researcher is in order!

We'll read in the data and take it from there:

::: {.panel-tabset group="language"}
## R

```{r}
messy_data <- read_csv("data/messy_data.csv")
```


## Python

```{python}
messy_data = pd.read_csv("data/messy_data.csv")
```

:::

### Variable naming {#sec-variable_naming}

The `messy` data set has many different issues with it. One of the issues is that the column headers are not consistent. This is an issue that you'll come across on a regular basis, since people have their own style (usually an inconsistent one!).

In both R and Python we have access to a "janitor" package ([`janitor`](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html) in R and [`pyjanitor`](https://pypi.org/project/pyjanitor/) in Python). These are fantastic packages that can save you tons of time. 
Here we'll just use one of the functions: `clean_names()`.

First, let's see what we're dealing with.

::: {.panel-tabset group="language"}
## R

```{r}
messy_data |> 
  colnames()
```

## Python

```{python}
messy_data.columns.tolist()
```

:::

That's not ideal. There is inconsistency in the use of capitalisation and there are full stops in the column names.

Now, let's see what the `clean_names()` function makes of all of this (without updating the names just yet).

::: {.panel-tabset group="language"}
## R

```{r}
messy_data |> 
  clean_names() |> 
  colnames()
```

## Python

```{python}
(messy_data           # data
  .clean_names()      # clean column names
  .columns.tolist())  # put column names in list
```

:::

We can see that the column names are now consistently lowercase and that the full stop `.` in the names have been replaced with `_` underscores.

I'm happy with that, so I'll assign those new names to the data.

::: {.panel-tabset group="language"}
## R

```{r}
messy_data <- messy_data |>
  clean_names()
```


## Python

```{python}
messy_data = messy_data.clean_names()
```

Sometimes we also want to include generic prefixes to our column names. We can do this with f-strings.

::: {.callout-note}
## Formatted string literals (f-strings)

These let you insert variables or expressions directly into a string by prefixing it with `f` or `F`.

Have a look at this example:

```{python}
course = "Data analysis in"
language = "Python"

message = f"This is the {course} in {language} course."

print(message)
```

We already used f-strings earlier in @sec-long_to_wide. There we created a prefix `yr_` for each column.

```{python}
#| eval: false
surveys_wide.columns = [f"yr_{col}" for col in surveys_wide.columns]
```

This example uses list comprehension: `[f"yr_{col}" for col in surveys_wide.columns]`. It loops through each `col` in `surveys_wide.columns` and for each column it creates an f-string `f"yr_{col}`. The value of `col` gets taken from whatever the column name is, in our case a number representing the year.
:::

:::

### Adjusting numbers

In the [joining section](#setup_joins) we saw that it wasn't great practice to just use numbers to indicate `plot_id`, since they obviously have no numerical value, but instead define a category. This is something that often occurs, so it's useful to know how to be able to format them differently.

For example, we could encode them in the format `plot_xxx` where `xxx` is a number with leading zeros (so that it sorts nicely).

We can do that as follows:

::: {.panel-tabset group="language"}
## R

```{r}
plot_types |> 
  mutate(plot_id = paste0("plot_", sprintf("%03d", plot_id)))
```

::: {.callout-tip collapse="true"}
## Alternative: using the `stringr` package

The `stringr` package is part of `tidyverse` and has a whole range of functions that allow you to change strings (text). The equivalent of the code above would be:

```{r}
plot_types |> 
  mutate(plot_id = str_c("plot_", str_pad(plot_id, width = 3, pad = "0")))
```

You can read this as: "Use `mutate()` to update the `plot_id` column, where (`=`) the contents of `plot_id` is a combined string (`str_c`) containing the text `"plot_` and we pad out (`str_pad`) `plot_id` so it's 3 digits wide (`width = 3`) and we pad out with `"0"` (`pad = "0"`)."

:::

If you wanted to update the column, just add `plot_types <-` in front of this!

## Python

```{python}
plot_types["plot_id"].apply(
    lambda x: f"plot_{x:03d}"
)
```

::: {.callout-tip collapse="true"}
## Alternative: using `.zfill()`

If you're not keen on this method using the `lambda x:` approach, you can use `.zfill`:

```{python}
"plot_" + plot_types["plot_id"].astype(str).str.zfill(3)
```

You can read this as: "Take the `plot_id` column from `plot_types` (`plot_types["plot_id"]`) and update it (`=`) with a combination of the text `plot_` (`"plot_"`) and the value of `plot_types["plot_id"]`, which needs to be converted to a string (`.astype(str)`) before we can fill it with zeros up to a maximum of 3 digits (`.str.zfill(3)`)."
:::

If you wanted to update the column, just add `plot_types["plot_id"] =` in front of this!

:::

Note: this means that you would also have to change the `plot_id` column values in the `surveys` data set, if you wanted to combine the data from these tables!

### Encoding issues

In the `messy_data` data set it's not just the column names that are an issue. There are quite a few different encoding issues. We will address several of them now, but some of these you'll investigate later in the exercises.

| Variable          | Problem(s)                                                                 |
|-------------------|----------------------------------------------------------------------------|
| `id`              | Clean — serves as a unique identifier                                      |
| `age`             | (to investigate later) |
| `gender`          | (to investigate later) |
| `score`           | Numeric-like variable contains text entries: `"five"`, `"high"`, `NA`      |
| `income_gbp`      | (to investigate later)                   |
| `country`         | Inconsistent naming for UK: `"UK"`, `"U.K."`, `"United Kingdom"`, `"United kingdom"`, `NA` |
| `employed_or_not` | Inconsistent boolean values: `"yes"`, `"y"`, `"TRUE"`, `"n"`, `NA`, etc. |
| `notes`           | Free-text notes with mixed missing indicators: `"N/A"`, `""`, `"none"`, `NA` |

Let's focus on the `country` column, which is a classical case of "different people have different ways of encoding the same thing".

First, we check what entries we have in this column. Apart from just showing the different types of entries, we're also looking at how many times they occur.

::: {.panel-tabset group="language"}
## R

```{r}
messy_data |> 
  count(country)
```

## Python

```{python}
counts = (messy_data
          .groupby("country")
          .size()
          .reset_index(name = "n"))

counts
```

:::

We've got quite some variation going on here. A convenient way of addressing this is to create a list of substitutions and then apply that to the data. This makes it easier in the future if you get more variations that you need to update.

::: {.panel-tabset group="language"}
## R

In R we have the `case_when()` function, which can help you do exactly that. It's more general though: it allows you to re-encode values based on conditions. So, before we apply it to our data, let's go through a simplified example.

::: {.callout-tip}
## Using `case_when()` for recoding values
It works in this way:

```{r}
# 1. create some dummy data
df <- tibble(score = c(45, 67, 82, 90, 55))

# 2. recode the values
df <- df |> 
  mutate(performance = case_when(
    score < 60               ~ "fail",
    score >= 60 & score < 80 ~ "pass",
    TRUE                     ~ "excellent"
  ))

# 3. show the result
df
```

In the mini-example above we create a simple data set with one column: `score`. There are 5 values in this column. We now want to assign a `performance` value to it, which will differ based on the score.

The `case_when()` function goes through each possible defined comparison (e.g. `score < 60`, `...`) and then assigns the value based on that (`~  "fail"`). We put the whole thing in a `mutate()` because we are creating a new column: `performance`.

At the end of the `case_when()` we have `TRUE ~ "excellent`. The `TRUE ~` designation means: "for everything else do...".

:::

Let's apply this to our `messy_data`.

```{r}
messy_data <- messy_data |> 
  mutate(country = case_when(
    country %in% c("United kingdom", "U.K.", "UK") ~ "United Kingdom",
    TRUE ~ country
  ))
```

1. `country = `indicates that we are updating our `country` column.
2. `country %in% c("United kingdom", "U.K.", "UK")` says, for each value in `country`, compare it to the values within `c()`
3. `~ "United Kingdom"` if it finds it, replace it with `"United Kingdom"
4. `TRUE ~ country` otherwise, leave it unchanged (including the missing values)

## Python

Let's apply this to our `messy_data`.

First we create a "translation" table: which values do we want to re-encode and how?

```{python}
# Define standard mapping for inconsistent names
country_map = {
    "United kingdom": "United Kingdom",
    "U.K.": "United Kingdom",
    "UK": "United Kingdom"
    # Add more variations if needed
}
```

Next, we apply this to our data:

```{python}
# Update the country names
messy_data["country"] = (
    messy_data["country"]
    .replace(country_map))
```
:::

We can view the end result:

::: {.panel-tabset group="language"}
## R

```{r}
messy_data |> 
  count(country)
```

## Python

```{python}
counts = (messy_data
         .groupby("country")
         .size()
         .reset_index(name = "n"))

counts
```

:::

### Boolean values
We have a column `employed_or_not`, which contains information on the employment status of each person. This is encoded inconsistently, as `"yes`, `"y"`, `"TRUE"`, ...

Apart from being consistent, it's useful to keep columns that can only have 1 of 3 value (true, false, missing) as a boolean. This makes it easier to filter and tally contents than if we'd encode it as text.

Let's tackle this issue in a similar approach as above. First, let's see what we're dealing with here, by looking up the column types and the contents of `employed_or_not`.

::: {.panel-tabset group="language"}
## R

```{r}
class(messy_data$employed_or_not)
```

```{r}
messy_data |>
  count(employed_or_not)
```

We can see that the column is viewed as a `"chr"` or character column.

## Python

```{python}
messy_data["employed_or_not"].dtype
```

We can see that the column is viewed as an `object` type (`'O'`). This type usually holds text, but can also be used for a mixture of Python objects.

Let's see what kind of values we have in our column:

```{python}
counts = (messy_data
          .groupby("employed_or_not", dropna = False)
          .size()
          .reset_index(name = "n"))

counts
```

:::

The reason why we end up with a generic data type is because there is a mixture of data in the column. Let's fixed that.


::: {.panel-tabset group="language"}
## R

```{r}
messy_data <- messy_data |> 
  mutate(employed_or_not = case_when(
    employed_or_not %in% c("no", "n") ~ "FALSE",
    employed_or_not %in% c("yes", "y") ~ "TRUE",
    TRUE ~ employed_or_not
  ))
```

Let's count again:

```{r}
messy_data |> 
  count(employed_or_not)
```

But the `employed_or_not` column is still viewed as `character`:

```{r}
class(messy_data$employed_or_not)
```

So, we need to force it to view it as such.

To make it a boolean / logical column, we use the `as.logical()` function.

```{r}
messy_data <- messy_data |> 
  mutate(employed_or_not = as.logical(employed_or_not))
```

We check one last time:

```{r}
class(messy_data$employed_or_not)
```

## Python

```{python}
# Define standard mapping for inconsistent names
employment_map = {
    "no": "False",
    "n": "False",
    "yes": "True",
    "y": "True"
    # Add more variations if needed
}
```

Next, we apply this to our data:

```{python}
# Update the country names
messy_data["employed_or_not"] = (
    messy_data["employed_or_not"]
    .replace(employment_map))
```

We count again to check what has happened:

```{python}
counts = (
    messy_data
    .groupby("employed_or_not", dropna = False)
    .size()
    .reset_index(name = "n")
)

counts
```

This still shows a difference between `FALSE` and `False`, and `TRUE` and `True`. That's because the all-caps words are interpreted as text. In fact, *all* of them are viewed as text, because we used `" "` in our mapping.

Python's boolean phrases are `True` and `False`, so we need to adjust that.

To do this, we first convert all the values to lower case with `.str.lower()`, and then convert those values to boolean (without `" "`). Lastly, we convert the column to `"boolean"`. We need to do that last step, because we have missing data. If we didn't, then Python would keep all the contents as text.

```{python}
messy_data["employed_or_not"] = (
    messy_data["employed_or_not"]
    .str.lower()
    .map({"true": True, "false": False})
    .astype("boolean")     # convert to boolean type
)

```

We can see that we only have `True`, `False` or `<NA>`.

```{python}
counts = (
    messy_data
    .groupby("employed_or_not", dropna = False)
    .size()
    .reset_index(name = "n")
)

counts
```

And our data type is now correct.

```{python}
messy_data["employed_or_not"].dtype
```

:::

Success!

## Cleaning plots

Often we need to put a bit of work into making plots more presentable. Perhaps default colours are not quite right, or labels need adjusting. Below we cover some common changes.

Let's use the `surveys_left` data set for the next few sections.If you haven't done so already, please load the following:

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: false
surveys <- read_csv("data/surveys.csv")
plot_types <- read_csv("data/plots.csv")
surveys_left <- left_join(surveys, plot_types, by = "plot_id")
```

## Python

```{python}
#| eval: false
surveys = pd.read_csv("data/surveys.csv")
plot_types = pd.read_csv("data/plots.csv")
surveys_left = pd.merge(surveys, plot_types, how = "left", on = "plot_id")
```

:::

### Adding titles and axis labels

First we visualise some data. Here we are creating a [violin plot](https://en.wikipedia.org/wiki/Violin_plot). Partly because we can, but mostly because they are useful. They can tell you a lot about how your data are distributed.

::: {.panel-tabset group="language"}
## R

```{r}
#| label: fig-violin_plottype
#| fig-cap: "Violin plot of hindfoot_length for each plot type"
ggplot(surveys_left, aes(x = plot_type, y = hindfoot_length)) +
         geom_violin()
```


## Python

```{python}
#| results: hide
#| label: fig-violin_plottype_py
#| fig-cap: "Violin plot of hindfoot_length for each plot type"
p = (ggplot(surveys_left, aes(x = "plot_type", y = "hindfoot_length")) +
         geom_violin())

p.show()
```

:::

It'd be nice to have **plot title**, as well as some clearer (and nicer) **axis labels**. We can do this as follows:

::: {.panel-tabset group="language"}
## R

```{r}
#| warning: false
#| label: fig-violin_plottype_title
#| fig-cap: "Violin plot of hindfoot_length for each plot type"
ggplot(surveys_left, aes(x = plot_type, y = hindfoot_length)) +
         geom_violin() +
  labs(title = "Violin plot for hindfoot length by plot type",
       subtitle = "1977 - 2022",
       x = "Plot type",
       y = "Hindfoot length (mm)")
```

## Python

```{python}
#| results: hide
#| label: fig-violin_plottype_title_py
#| fig-cap: "Violin plot of hindfoot_length for each plot type"
p = (ggplot(surveys_left, aes(x = "plot_type", y = "hindfoot_length")) +
         geom_violin() +
       labs(title = "Violin plot for hindfoot length by plot type",
            subtitle = "1977 - 2022",
            x = "Plot type",
            y = "Hindfoot length (mm)"))

p.show()
```

:::

### Text axis labels: rotate

Another common issue is that axis text labels are poorly aligned. In the plots above the `plot_type` values are too long and the text overlaps. Let's fix this. We'll do this in two ways:

1. Rotating the axis text
2. Wrapping / inserting a line break in the axis text

::: {.panel-tabset group="language"}
## R

```{r}
#| warning: false
#| label: fig-violin_plottype_textrotate
#| fig-cap: "Rotating axis text labels."
ggplot(surveys_left, aes(x = plot_type, y = hindfoot_length)) +
  geom_violin() +
  labs(title = "Violin plot for hindfoot length by plot type",
       subtitle = "1977 - 2022",
       x = "Plot type",
       y = "Hindfoot length (mm)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Python

```{python}
#| results: hide
#| label: fig-violin_plottype_textrotate_py
#| fig-cap: "Rotating axis text labels."
p = (ggplot(surveys_left, aes(x = "plot_type", y = "hindfoot_length")) +
         geom_violin() +
       labs(title = "Violin plot for hindfoot length by plot type",
            subtitle = "1977 - 2022",
            x = "Plot type",
            y = "Hindfoot length (mm)") +
        theme(axis_text_x = element_text(rotation = 45, ha = "right")))

p.show()
```


:::

### Text axis labels: wrap

Rotating the text labels can work in a lot of cases, but doesn't get round the issue of long strings. The `plot_type` values aren't exactly concise. So, it might be better to have some text wrapping instead.

::: {.panel-tabset group="language"}
## R

```{r}
#| warning: false
#| label: fig-violin_plottype_textwrap
#| fig-cap: "Wrapping axis text labels."
ggplot(surveys_left, aes(x = plot_type, y = hindfoot_length)) +
  geom_violin() +
  labs(
    title = "Violin plot for hindfoot length by plot type",
    subtitle = "1977 - 2022",
    x = "Plot type",
    y = "Hindfoot length (mm)") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))
```

## Python

Python has a packages called `textwrap` that can help here.

```{python}
import textwrap
```

Although the code is a bit convoluted (it's a case of copy/paste in future plots, really) it does result in a much cleaner plot:

```{python}
#| results: hide
#| label: fig-violin_plottype_textwrap_py
#| fig-cap: "Wrapping axis text labels."
p = (ggplot(surveys_left, aes(x = "plot_type", y = "hindfoot_length")) +
         geom_violin() +
       labs(title = "Violin plot for hindfoot length by plot type",
            subtitle = "1977 - 2022",
            x = "Plot type",
            y = "Hindfoot length (mm)") +
       scale_x_discrete(labels = lambda labels: [
                          textwrap.fill(str(l), 10) if l is not None else ""
                          for l in labels]))

p.show()
```

:::

## Changing colours

Often we want to use colours to emphasise parts of our plot or to highlight different groups within our data. Default options might not always be the most appropriate ones, so it's good to know how to control the aesthetics of your plot. We'll cover that below. We'll use a small subset of the data so we can highlight the differences better.

::: {.panel-tabset group="language"}
## R

```{r}
surveys_y2k <- surveys_left |> 
  filter(plot_type %in% c("Control", "Rodent Exclosure")) |> 
  filter(year %in% c(2000, 2001, 2002)) |> 
  drop_na()
```

## Python

```{python}
surveys_y2k = (
    surveys_left[
        surveys_left["plot_type"].isin(["Control", "Rodent Exclosure"])
        & surveys_left["year"].isin([2000, 2001, 2002])]
    .dropna())
```

:::

### Colours: default

The default here is various shades of grey.

::: {.panel-tabset group="language"}
## R

```{r}
#| label: fig-scatbox_default
#| fig-cap: "Boxplot and jittered points, default."
ggplot(surveys_y2k, aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  geom_jitter(width = 0.1)
```

## Python

```{python}
#| results: hide
#| label: fig-scatbox_default_py
#| fig-cap: "Boxplot and jittered points, default."
p = (ggplot(surveys_y2k, aes(x = "plot_type", y = "hindfoot_length")) +
  geom_boxplot() +
  geom_jitter(width = 0.1))
  
p.show()
```

:::

Let's colour some of our data using the information from the `sex` variable.

::: {.panel-tabset group="language"}
## R

```{r}
#| label: fig-scatbox_bysex
#| fig-cap: "Boxplot and jittered points, default colours."
ggplot(surveys_y2k, aes(x = plot_type, y = hindfoot_length, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1)
```

## Python

```{python}
#| results: hide
#| label: fig-scatbox_bysex_py
#| fig-cap: "Boxplot and jittered points, default colours."
p = (ggplot(surveys_y2k,
            aes(x = "plot_type", y = "hindfoot_length", fill = "sex")) +
  geom_boxplot() +
  geom_jitter(width = 0.1, fill = "black"))
  
p.show()
```
:::

We can see that some default colours get applied. Let's update this with some other default colour scheme.

::: {.callout-note collapse=true}
## Colour inheritance in `ggplot2` vs `plotnine`

You might have noticed that we've specifically given the instruction to `geom_jitter()` to use black: `geom_jitter(width = 0.1, fill = "black"))`. This is because colour inheritance in `plotnine` works subtly different from `ggplot2`.

In `ggplot2`, `geom_jitter()` by default uses the `colour` aesthetic (outline color) and does not inherit `fill` (which is used for boxplots and polygons). This is why points appear black in R. However, `plotnine` is less consistent and therefor you need to specify the colour yourself, otherwise the `geom_jitter()` data would also get coloured by `sex` in the example above.
:::

### Colours: standard palettes

A popular colour scheme is the so-called **Brewer palette**. Although [originally developed for R](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html), the colour combinations are appropriate in many contexts. It has colour palettes that are appropriate for data continuous data or ordered data, where colour signifies a difference in magnitude. It also contains colour palettes for qualitative data, where the emphasis is on creating visual separation between different groups.

::: {.panel-tabset group="language"}
## R

The `brewer` set of colours can be called using a set of functions in `ggplot2`. We'll mostly use two of these:

* `scale_fill_brewer()` for objects in a plot that have a surface area (such as boxplots)
* `scale_colour_brewer()` for objects that do not have a surface area (such as points)

We simply define the one we want, then add the type of palette with `palette =`. Here we use `"Dark2"`, which is particularly suited for clear separation between categories.

```{r}
#| label: fig-scatbox_bysex_brewer
#| fig-cap: "Boxplot and jittered points, Brewer (Dark2) colours."
ggplot(surveys_y2k, aes(x = plot_type, y = hindfoot_length, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_fill_brewer(palette = "Dark2")
```

## Python

The `brewer` set of colours can be called using a set of functions in `plotnine`. We'll mostly use two of these:

* `scale_fill_brewer()` for objects in a plot that have a surface area (such as boxplots)
* `scale_colour_brewer()` for objects that do not have a surface area (such as points)

We simply define the one we want, then add the type of palette with `palette =`. Here we use `"Dark2"`, which is particularly suited for clear separation between categories. We also need to define that the data are qualitative, using `type = qual`.

```{python}
#| results: hide
#| label: fig-scatbox_bysex_brewer_py
#| fig-cap: "Boxplot and jittered points, Brewer (Dark2) colours."
p = (ggplot(surveys_y2k,
            aes(x = "plot_type", y = "hindfoot_length", fill = "sex")) +
  geom_boxplot() +
  geom_jitter(width = 0.1, fill = "black") +
  scale_fill_brewer(type = "qual", palette = "Dark2"))
  
p.show()
```
:::

### Colours: colour-blind friendly {#cb_friendly}

Colour blindness is a reasonably common occurrence, with around 4.5% of the U.K. population having some form, according to @colourblindawareness.

The `brewer` palettes often have a good visual separation, but might not always be the most optimal ones. There is a set of colour-blind friendly colours developed by @okabeandito.

```{r}
#| echo: false
cb_palette   <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                  "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#| echo: false
#| label: fig-okabeandito
#| fig-cap: "Colour-blind friendly colour palette by @okabeandito."
scales::show_col(cb_palette)
```

If we want to apply these custom colours, we'll have to do that manually. The easiest way is to assign them to a manual colour palette, and then use that when plotting.

::: {.panel-tabset group="language"}
## R

```{r}
cb_palette   <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                  "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#| label: fig-scatbox_bysex_cb
#| fig-cap: "Boxplot and jittered points, Okabe and Ito colours."
ggplot(surveys_y2k, aes(x = plot_type, y = hindfoot_length, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_fill_manual(values = cb_palette)
```

## Python

```{python}
cb_palette = ["#999999", "#E69F00", "#56B4E9", "#009E73",
              "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]
```

```{python}
#| results: hide
#| label: fig-scatbox_bysex_cb_py
#| fig-cap: "Boxplot and jittered points, Okabe and Ito colours."
p = (ggplot(surveys_y2k,
            aes(x = "plot_type", y = "hindfoot_length", fill = "sex")) +
  geom_boxplot() +
  geom_jitter(width = 0.1, fill = "black") +
  scale_fill_manual(values = cb_palette))
  
p.show()
```
:::

## Arranging plots

We've already seen that we can use facetting to divide plots into sub-plots. Sometimes it can be very useful to combine completely separate plots into a single figure. We'll show how to do this below.

Let's keep using our `surveys_y2k` data set. We'll first create three plots (randomly chosen) to play around with. Just copy/paste the code below.

::: {.panel-tabset group="language"}
## R

```{r}
p1 <- ggplot(surveys_y2k, aes(x = plot_type, y = weight)) +
  geom_boxplot() +
  labs(title = "plot 1")

p2 <- ggplot(surveys_y2k, aes(x = weight, y = hindfoot_length)) +
  geom_point() +
  labs(title = "plot 2")

p3 <- ggplot(surveys_y2k, aes(x = factor(month), y = weight)) +
  geom_boxplot() +
  labs(title = "plot 3")
```


## Python

```{python}
#| results: hide

p1 = (ggplot(surveys_y2k, aes(x='plot_type', y='weight')) +
      geom_boxplot() +
      labs(title="plot 1"))

p2 = (ggplot(surveys_y2k, aes(x='weight', y='hindfoot_length')) +
      geom_point() +
      labs(title="plot 2"))

p3 = (ggplot(surveys_y2k, aes(x='factor(month)', y='weight')) +
      geom_boxplot() +
      labs(title="plot 3"))
```

:::

::: {.panel-tabset group="language"}
## R

In R we can use the `patchwork` package. Install it, if needed, with:

```{r}
#| eval: false
install.packages("patchwork")
```

And then load it:

```{r}
library(patchwork)
```

::: {.callout-note}
We're not covering the functionality of `patchwork` extensively here, since there is a very good [vignette](https://patchwork.data-imaginist.com) doing that. So, for more details, please refer to that link!
:::

## Python

In Python there is a `patchworklib` that tries to emulate what R's `patchwork` is doing. However, it seems to run into issues quite quickly when `plotnine` gets updated.

So, we don't provide code for this for the time being. We'll revisit this once the package is more stable.

:::


We can easily combine the 3 plots we created into a single panel.

::: {.panel-tabset group="language"}
## R

```{r}
(p1 + p2) / p3
```


## Python

:::

Often it's helpful to add labels, either to the figure itself and/or to the subpanels.

::: {.panel-tabset group="language"}
## R

We can add a title using the `title =` and add tags using `tag_levels =` arguments. These get added to the plot, using `plot_annotation()`:

```{r}
(p1 + p2) / p3 +
  plot_annotation(title = "Fig. 1: Trialling composite plots",
                  tag_levels = "a")
```

You can change the `tag_levels =` value to `"A"` for uppercase tags, or `"1"` for numbers (which wouldn't be very clear).

## Python

:::

## Summary

::: {.callout-tip}
#### Key points

::: {.panel-tabset group="language"}
## R

- We can use the `clean_names()` function from the `janitor` package to clean up column names, avoiding spaces, inconsistent case use etc.
- Try to avoid ID columns with numbers only, but instead use a prefix, so the data are not viewed as numerical.
- Common encoding issues include: inconsistent (upper) case use (e.g. `Belgium` vs `belgium`); poorly defined boolean values (`y` / `yes` instead of `TRUE` or `FALSE`).
- It is good practice to include a plot `title` and well-defined `x` and `y` axis labels.
- Ensure axes labels are clear by wrapping text or, less ideal, rotating them.
- Use colours wisely, ensuring separation between groups.
- Consider colour blindness in your audience and use colour-blind safe colour schemes.
- Arrange plots using `patchwork`

## Python

- We can use the `clean_names()` function from the `pyjanitor` package to clean up column names, avoiding spaces, inconsistent case use etc.
- Try to avoid ID columns with numbers only, but instead use a prefix, so the data are not viewed as numerical.
- Common encoding issues include: inconsistent (upper) case use (e.g. `Belgium` vs `belgium`); poorly defined boolean values (`y` / `yes` instead of `True` or `False`).
- It is good practice to include a plot `title` and well-defined `x` and `y` axis labels.
- Ensure axes labels are clear by wrapping text or, less ideal, rotating them.
- Use colours wisely, ensuring separation between groups.
- Consider colour blindness in your audience and use colour-blind safe colour schemes.

:::
:::
