---
title: "Live demo exercises"
---

## Section setup

::: {.panel-tabset group="language"}
## R

```{r}
#| message: false
library(tidyverse)
```

## Python

```{python}
# A Python data analysis and manipulation tool
import pandas as pd

# Python equivalent of `ggplot2`
from plotnine import *

# If using seaborn for plotting
import seaborn as sns
import matplotlib.pyplot as plt
```

:::

## DA1: Getting started

## DA2: Data & plotting

### Live demo: exploring `infections`

::: {.panel-tabset group="language"}
## R

Read in the data:

```{r}
#| message: false
infections <- read_csv("data/infections.csv")
```

And have a look:

```{r}
#| eval: true
head(infections)
```


## Python

Read in the data:

```{python}
infections = pd.read_csv("data/infections.csv")
```

And have a look:

```{python}
#| eval: true
infections.head()
```

:::

#### Data structure

Number of rows & columns:

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: true
nrow(infections)
```

```{r}
#| eval: true
ncol(infections)
```

## Python

```{python}
#| eval: true
infections.shape[0]
```

```{python}
#| eval: true
infections.shape[1]
```


:::

It's good to look at the column attributes: what type of columns are we dealing with and is it what we expect?

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: true
summary(infections)
```

```{r}
#| eval: true
str(infections)
```

## Python

```{python}
#| eval: true
infections.describe()
```

:::

#### Quality control checks

It's good to do some basic sanity / quality control checks. For example, if there are different categories in a column, do all the categories we expect show up or are there missing ones / misspelled etc.?

For example, we can check the unique values in a column:

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: true
unique(infections$infection_type)
```

## Python

```{python}
#| eval: true
infections["infection_type"].unique()
```

:::

We can count the number of missing values in the column `infection_type`.

::: {.panel-tabset group="language"}
## R

You read the code "inside-out":

```{r}
#| eval: true
sum(is.na(infections$infection_type))
```

## Python

```{python}
#| eval: true
infections["infection_type"].isna().sum()
```

:::

### Live demo: subsetting `infections`

Let's select `patient_id`:

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: false
infections$patient_id
```

## Python

```{python}
#| eval: true
infections.patient_id
```

:::

Or more than 1 column, by column name:

::: {.panel-tabset group="language"}
## R

```{r}
infections[, c("patient_id", "systolic_pressure")]
```

## Python

```{python}
#| eval: true
infections[["patient_id", "systolic_pressure"]]
```

:::

Combine this with selecting only a subset of rows, let's say the first three rows.

::: {.panel-tabset group="language"}
## R

```{r}
#| eval: true
infections[1:3, c("patient_id", "systolic_pressure")]
```

## Python

We need to be aware of the zero-based indexing, also noting that the value after the `:` is *not* included:

```{python}
#| eval: true
infections[["patient_id", "systolic_pressure"]].iloc[0:3]
```

:::

### Live demo: simple plot `infections`

Let's start with a simple scatterplot, where we plot `body_temperature` on the x-axis and `crp_level` on the y-axis.

::: {.panel-tabset group="language"}
## R

```{r}
ggplot(infections, aes(x = body_temperature, y = crp_level)) +
  geom_point()
```

## Python

```{python}
#| results: hide

p = (ggplot(infections, aes(x = "body_temperature", y = "crp_level")) +
  geom_point())
  
p.show()
```

:::

We can explore this a bit further. For example, we can colour the points based on the `hospital` variable, to see if there are any patterns across the different hospitals:

::: {.panel-tabset group="language"}
## R

```{r}
ggplot(infections, aes(x = body_temperature, y = crp_level,
                       colour = hospital)) +
  geom_point()
```

## Python

```{python}
#| results: hide

p = (ggplot(infections, aes(x = "body_temperature", y = "crp_level",
                            colour = "hospital")) +
  geom_point())
  
p.show()
```

:::

### Live demo: facetting `infections`

When we plotted `body_temperature` and `crp_level` against each other and coloured the data based on `hospital`, we ended up with a rather unclear plot. This is probably because there are no clear differences between the hospitals. However, we can separate these data a bit more clearly by using facets.

::: {.panel-tabset group="language"}
## R

```{r}
ggplot(infections, aes(x = body_temperature, y = crp_level,
                       colour = hospital)) +
  geom_point() +
  facet_wrap(facets = vars(hospital))
```

## Python

```{python}
#| results: hide

p = (ggplot(infections, aes(x = "body_temperature", y = "crp_level",
                            colour = "hospital")) +
  geom_point() +
  facet_wrap("~ hospital"))
  
p.show()
```

:::

## DA3: Manipulating data

## DA4: Organise and combine

## Live demo: joining `hospital_info`

In the `infections` data set we have a variable called `hospital`. This contains the following unique entries:

::: {.panel-tabset group="language"}
## R

```{r}
infections |> count(hospital)
```

## Python
:::

Note that for some of the `hospital` entries there are missing data. This is relevant later on, when we're joining.

We're now going to add information on the hospitals, which are stored in `hospital_info.csv`. Let's read in the data:

::: {.panel-tabset group="language"}
## R

```{r}
#| message: false
hospital_info <- read_csv("data/hospital_info.csv")
```


## Python
:::

Look at the data. There are `r hospital_info |> distinct(hospital) |> nrow()` distinct hospital entries, with additional information for each hospital. We have the following variables:

| Variable           | Type        | Description                                                                 |
|--------------------|-------------|-----------------------------------------------------------------------------|
| `hospital`         | character (id) | Unique hospital identifier (`hospital_1` … `hospital_6`).                   |
| `hospital_name`    | character   | Official hospital name (e.g., *Royal London Hospital*).                     |
| `location`         | character   | City where the hospital is located (e.g., London, Manchester).              |
| `bed_capacity`     | integer     | Approximate number of inpatient beds available at the hospital.             |
| `teaching_hospital`| logical     | Indicates if the hospital is a teaching hospital (`TRUE` / `FALSE`).   |



### Left join

First, we'll add the `hospital_info` data to the `infections` data set. We do this with a left-join. We expect hospital info data to be added to our main `infections` table, if the `hospital` value in `infections` matches with the one in `hospital_info`.

::: {.panel-tabset group="language"}
## R

```{r}
infections_left <- left_join(infections, hospital_info, by = "hospital")
```

Let's just select the ID column from `infections`, together with all the columns that have been added.

```{r}
infections_left |> 
  select(patient_id, hospital, hospital_name, location, bed_capacity, teaching_hospital) |> 
  head()
```

We can see that the hospital information is now added to the data. We don't expect any data to have dropped, so the number of observations/rows in `infections_left` should match the original `r nrow(infections)` from `infections`.

```{r}
nrow(infections_left)
```

## Python

:::

### Right join

So, how does a right join then differ from a left join? Well, here we'd be adding the `infections` data to the `hospital_info` data (so, in the other direction). That means that for each `hospital` value that exists in `hospital_info` it will try and find the values that match the `hospital` column in `infections`.

::: {.panel-tabset group="language"}
## R

```{r}
infections_right <- right_join(infections, hospital_info, by = "hospital")
```

Again, the resulting table contains the data from both `infections` and `hospital_info`:

```{r}
head(infections_right)
```

How many observations do we expect? Let's have a look at how many we've got.

```{r}
nrow(infections_right)
```


## Python

:::

Here we see that we have fewer than the original `r nrow(infections)` observations, because we are joining in the other direction. Remember, there are missing values in our `infections` data set (`r sum(is.na(infections$hospital))` in total). When we're joining `infections` **to** `hospital_info`, these entries get dropped, because there is no missing value entry in the `hospital` column of `hospital_info`.

For the eagle-eyed amongst you: we have `r nrow(infections_right)` observations/rows in `infections_right`, whereas there are `r sum(is.na(infections$hospital))` missing values. We might have expected there to be `r nrow(infections) - sum(is.na(infections$hospital))` rows (the difference between the number of rows in `infections` and the number of missing values), but there are `r nrow(infections_right)`.

Why? Well, there is one entry in `hospital_info` that does not appear in `infections`: `hospital_6`. When we right join, this value gets retained, so this adds one additional row to the final output.

::: {.panel-tabset group="language"}
## R

```{r}
infections_right |> count(hospital)
```


## Python

:::

If we wouldn't want that, we'd use inner join.

### Inner join

Inner join will join two tables and only retain values (based on the joining ID) that exist in *both* tables.

::: {.panel-tabset group="language"}
## R

```{r}
infections_inner <- inner_join(infections, hospital_info, by = "hospital")
```

```{r}
nrow(infections_inner)
```

Only the 5 hospitals that appear in the `infections` data set are retained:

```{r}
infections_inner |> count(hospital)
```


## Python

:::

This is entirely expected, since we don't have missing data in `hospital_info` (so all the missing data from `infections` are dropped), nor do we have a `hospital_6` entry in `infections` (so that is dropped too).

### Full join

If we wanted to retain *all* observations, regardless of which table they're from, we'd use a full join.

::: {.panel-tabset group="language"}
## R

```{r}
infections_full <- full_join(infections, hospital_info, by = "hospital")
```

How many rows do we expect? Well, the `r nrow(infections)` from `infections` plus the one extra (`hospital_6` entry) from `hospital_info`:

```{r}
nrow(infections_full)
```

We can check that we've retained all values, by counting the number of observations for each `hospital` value:

```{r}
infections_full |> count(hospital)
```


## Python

:::

We see that **all** `hospital` values are retained, including the missing values *and* the one `hospital_6` value coming from the `hospital_info` data set. Success!
